\chosentitle

### 2024-09-23 paper/main.tex
word-smithing

We might simply consider what is the best continuum for every possible theoretical spectrum under comparison. When there are thousands of model spectra to compare with, then for each spectrum we can take the ratio of the data and the model, apply some averaging filter (e.g., moving mean or median) and take that as the pseudo-continuum. These kinds of approaches work remarkably well, despite the requirement of a set of (sufficiently) comparable model spectra, and the added computational expense.\\

@@

Here we will describe the method in general before outlining the implementation details. We assume the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$). The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
\begin{align}
    y_i &= f(\lambda_i;\vecalpha)\hadamard{}g(\lambda_i;\vecbeta)\hadamard{}h(\lambda_i;\vecgamma) + \mbox{noise}
\end{align}
where the components $f(...)$, $g(...)$, and $h(...)$ are defined below. Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
\begin{eqnarray}
    \vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + \frac{2\sigma_{y,i}^3}{8y_i^3} - \frac{6\sigma_{y,i}^4}{24y_i^4}\right)^2
\end{eqnarray}
\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into a convenient summation:
\begin{align}
    \label{eq:log_y}
    \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
\end{align}\\

@@

We refer to our $N \times D$ matrix of continuum-rectified theoretical stellar spectra as $\vec{S}$. This is a dense matrix: there are no entries of exactly zero, with many entries near 1 (no absorption). However, a small transformation to this matrix makes it extremely sparse. Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
\begin{eqnarray}
    \label{eq:nmf}
    -\log\left({\vec{S}}\right) \approx \vec{W}_\star\transpose\vec{F}
\end{eqnarray}
where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}. The factorization of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative. The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\%Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\

%on previous works on NMF and data analysis more broadly. We would like the naming conventions to be consistent with those papers, but this nomenclature is easily overloaded. For this reason, we make clear definitions: throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights} and $\vecH$ as the NMF \emph{basis vectors}. \\

%Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ basis weights per spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding basis vectors each with $D$ pixels.

%The nomenclature  draws on previous works on NMF and data analysis, but this nomenclature is easy to overload, so throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights}, $\vecalpha$ as the NMF \emph{basis coefficients}, $\vecH$ as the NMF \emph{basis vectors}, and $\vectheta$ (not yet defined) will refer to the amplitudes of the sine-and-cosine basis.

The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages. In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages. We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, \todo{until reaching a plateau in reconstruction error, at which point we would iterate further using multiplicative updates.} The multiplicative update rules are appealing because they are convex, do not require many choices, and they can be efficiently iterated on a graphics card.\\

There are very few requirements of the theoretical spectral grid. There are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing in between points. The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to fit. This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of A- to M-type stars was also sufficiently flexible to model many O-type stars. For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\

We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model. This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \project{ATLAS9} code with \project{MARCS} spherical and plane-parallel models atmospheres. While the grid does include spectra for hotter stars computed with ATLAS atmospheres, we restricted ourselves to MARCS model atmospheres for this work. The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum. We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$, truncated the convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements. The matrix $\vec{S}$ includes all \todo{X} convolved BOSZ spectra, with \todo{N} pixels each, which we randomly shuffled before factorization. We then factorized this matrix into $K = 16$ basis vectors by coordinate descent for \todo{10,000} steps, \todo{and then X steps using multiplicative updates}. After training, the Frobenius loss, also known as the L2 reconstruction loss
\begin{eqnarray}
    L_2 = \norm{{\left(-\log{\vec{S}}\right) - \vec{W}_\star\transpose\vec{F}}}^2
\end{eqnarray}
\noindent{}was found to be \todo{X}. Figure~\ref{fig:schematic} shows some example basis vectors computed by this factorization, where it is somewhat striking that the basis vectors are identifiable by astrophysical mechanism, since no stellar parameters (or astrophysics of any kind) are included in the factorization process. For example, one basis vector just includes hydrogen lines. Another basis vector only has CO-molecule lines. We expand on this observation in later sections.\\

@@

Telluric transmission can be included in a similar way to stellar absorption. We used a grid of model telluric transmission spectra computed for the \texttt{PypeIt} project. These spectra are computed using the Line-by-Line Radiative transfer Model for an array of observatory locations (latitude, longitude) and atmospheric temperatures, pressures, humidity, and airmass. The original grid of telluric transmission spectra was not available to us. Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, and compressed using Principal Component Analysis and an inverse hyperbolic sine transformation, to approximate the original grid. With this reconstructed grid, we factorized the negative log of theoretical telluric absorption
\begin{eqnarray}
    -\log\vec{M} \approx \vec{W}\transpose_\earth\vec{G}
\end{eqnarray}
\noindent{}into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$ using $L = 4$ basis vectors. We took \todo{10,000} iterations using coordinate descent, at which point the reconstruction loss was \todo{X}. With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
\begin{eqnarray}
    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta
\end{eqnarray}
where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} in to be solved at inference time.
%We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}).
Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\

@@

\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
\begin{equation}
    \vec{Y} = -\vec{F}\transpose\vecalpha - \vec{G}\transpose\vecbeta + \vec{H}\vecgamma
\end{equation}
\noindent{}which we can compactly write as
\begin{equation}
    \vec{Y} = \vec{A}\vec{X}
\end{equation}
where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
\begin{eqnarray}
    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\-\vec{G}\transpose\\\vec{H}\end{bmatrix}
    \quad \mbox{and} \quad
    \vec{X} = \begin{bmatrix}\vecalpha\\\vecbeta\\\vecgamma\end{bmatrix} \quad .
\end{eqnarray}

@@

For this reason, we queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from B- to M-type. This selection is not intended to be complete, but is meant to demonstrate the effectiveness of a single linear model for fitting stars across a range of stellar parameters and evolutionary states. For each example star we downloaded a pseudo-random exposure, the details of which are shown in Table~\ref{tab:stars}. No explicit cut was made to ensure a minimum S/N ratio, but in practice most exposures are of high quality. The B- and A-type example stars show significant rotational broadening. For these stars, we adopt the $v\sin{i}$ values reported by \citet{Someone} and convolve the stellar basis vectors before inference. In Section~\ref{sec:discussion} we discuss options when radial velocity or rotational broadening measurements are available.\\

@@

The final product of the \harps\ data reduction pipeline is a one-dimensional unrectified spectrum, resampled onto a single wavelength array where all echelle orders have been stitched together. Because we are interested in fitting the continuum on a per-order basis, we instead use the \texttt{e2ds} data products. These are extracted one-dimensional unrectified spectra of each echelle order, with the wavelength solution represented by a low order polynomial. In early experiments we fit the spect ra in these files directly, an example of which can be seen in Figure~\ref{fig:harps-alf-cen-a}. However, for some exposures we noticed systematic residuals that were consistent with a 2D artefact in the image plane (e.g., a ghost or reflection). While we expect that flat-fielding should have already removed these kinds of artefacts, it did prompt us to explore using the nightly blaze calibration files. Using the calibration blaze files mitigated much of the visible structure, and expectedly provided spectra with a simpler, flatter shape. For these reasons, in all experiments except for the one shown in Figure~\ref{fig:harps-alf-cen-a}, we chose to fit blaze-corrected echelle orders.\\

When fitting each exposure we took the computed Barycentric Earth radial velocity and the radial velocity measured by the \harps\ data reduction pipeline to shift the stellar absorption basis vectors $\vec{F}$ to the observed frame. In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citep{Someone}. Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and \todo{9} Fourier modes per echelle order, totalling \todo{668} model parameters.\\

We used the same model setup for the second experiment, where we required many spectra taken of the same star at low and high (e.g., 10 to 200) S/N ratios. The lower bound on S/N ratio might be unrealistic since spectroscopists prefer measuring stellar parameters and chemical abundances from higher quality spectra, but the bound is necessary to evaluate performance at low S/N ratios. We searched the \eso/\harps/ catalog and found that not many stars meet this requirement. When sorted by number of exposures in descending order, Alpha Centauri A was the first star that met our requirements in repeated S/N sampling. We constructed bins of SNR from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum (Table~\ref{tab:obs-alf-cen-a}) for every SNR bin interval to provide an even sampling of spectra at low and high S/N ratios. We fit our model to each spectrum, treating it as totally indepdendent from all others.\\

\begin{figure*}
    \includegraphics[width=\textwidth, height=\textwidth]{../code/harps_examples.pdf}
    \caption{Continuum-rectified \harps\ spectra of example spectral types from B to M. The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission.\label{fig:harps-examples}}
\end{figure*}

@@

The first defined experiment involves fitting stars of a variety of spectral types. The rectified spectra for selected example stars in this experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type. From these rectified spectra it is clear that the model performs well on a variety of stellar types and evolutionary stages. HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorized using NMF. The fit to HD 45418 is quite good, despite this star being well outside the model grid. All other example stars shown in Figure~\ref{fig:harps-examples} have literature stellar parameters that place them well within the bounds of the factorized model grid. Most stars observed with \eso/\harps\ are predictably on the main-sequence, since they are more frequently subject to radial velocity exoplanet searches than evolved stars. HD 203050 (K2) is the only red giant branch in our sample. This is not obvious from Figure~\ref{fig:harps-examples}, suggesting the model does just as well on red giant branch stars as it does on main-sequence stars. This has been our experience in other works using this method on SDSS-V optical and near-infrared spectra (Casey et al., in prep).\\

The cooler stars in the example set have almost no so-called `continuum pixels' blueward of 4000\,\AA. This is expected; the continuum estimate provided by the model matches theoretical rectified spectra. It illustrates that the continuum estimate that we fit simultaneously with stellar absorption does provide a closer estimate to the true continuum (multiplied by the instrument response function) rather than a pseudo-continuum.\\

There is good agreement between the model fits and data for example stars of different spectral types. In detail there are mis-matches in line absorption strength, or mis-matches in the form of missing lines (in either the model or the data). We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both. In any case, the model appears to be sufficiently accurate to model the continuum without substantial bias, so it is not a major concern for our purposes. A separate reason for model-data disagreement comes from our use of the telluric transmission model. The telluric models available to us are of lower spectral resolution ($\mathcal{R} \sim 25{,}000$) than \harps. This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:alf-cen-A-example}), but it is otherwise not obvious. While the model telluric transmission is broader than the data, the telluric basis vectors are sufficiently close to the data that the telluric features in the data are well-captured by the model telluric basis vectors, and they do not adversely affect the continuum fits.\\

% TODO: anything else to say results-wise about experiment 1? \chi^2 values? systematic differences as a function of flux?

The second defined experiment tests how consistently we can predict continuum for the same star, as a function of S/N ratios. We fit \todo{X} spectra of Alpha Centauri A, which are sampled uniformly in S/N bins from 10 to 200 pixel$^{-1}$. We are now forced to select a statistic to evaluate the consistency of our continuum rectification. We adopt the 95th percentile in rectified flux as a statistic to compare to normalisation methods described in the literature, but we prefer to view how the \emph{distribution} of normalized flux values change as a function of S/N ratio. Both properties are shown in Figure~\ref{fig:repeatability-with-snr}, where the line indicates the 95th percentile of normalized flux, and the shaded area shows the distribution of normalized flux values. Above S/N $\sim$ 30, both indicators are extremely stable and do not vary significantly with S/N ratio. This is notable, as most spectra acquired for the purpose of stellar parameters and chemical abundances will aim for S/N $>$ 30, but spectra below S/N $<$ 50 are routinely described as being problematic to rectify consistently. From this figure it highlights that we can achieve consistent continuum normalisation whether the star has S/N $\sim$ 30 or 300.\\

We can make quantitative statements on how consistent the continuum rectification is. Taking the 95th percentile of normalized flux as a proxy statistic for the normalization consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalized flux is 0.22\% (a variation in normalized flux of 0.0022). This variation is smaller than the typical flux errors. This statistic increases only marginally when we include lower quality spectra. Taking all spectra with S/N > 50, we find the scatter in the 95th percentile of normalized flux to be 0.35\%. With S/N > 30, this rises marginally to 0.46\%. \todo{In all scenarios the error in continuum is less than the typical flux error.}\\

>>>>

We might simply consider what is the best continuum for every possible theoretical spectrum under comparison. When there are thousands of model spectra to compare with, then for each spectrum we can take the ratio of the data and the model, apply some averaging filter (e.g., moving mean or median) and take that as the pseudo-continuum. These kinds of approaches work remarkably well, despite the requirement of a (potentially very large) set of sufficiently similar model spectra, and the added computational expense.\\

@@

Here we will describe the method in general before outlining the implementation details. We assume the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$). The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
\begin{align}\label{eq:y}
    y_i &= f(\lambda_i;\vecalpha)\hadamard{}g(\lambda_i;\vecbeta)\hadamard{}h(\lambda_i;\vecgamma) + \mbox{noise}
\end{align}
where the components $f(...)$, $g(...)$, and $h(...)$ are defined below. Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
\begin{eqnarray}
    \vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + \frac{2\sigma_{y,i}^3}{8y_i^3} - \frac{6\sigma_{y,i}^4}{24y_i^4}\right)^2
\end{eqnarray}
\noindent{}which is computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
\begin{align}
    \label{eq:log_y}
    \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
\end{align}\\

@@

We refer to our $N \times D$ matrix of continuum-rectified theoretical stellar spectra as $\vec{S}$. This is a dense matrix: there are no entries of exactly zero, with many entries near 1 (no absorption). However, a small transformation to this matrix makes it extremely sparse. Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
\begin{eqnarray}
    \label{eq:nmf}
    -\log\left({\vec{S}}\right) \approx \vec{W}_\star\transpose\vec{F}
\end{eqnarray}
where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}. The factorization of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative. The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\

The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages. In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages. We found very similar looking basis vectors if we initialized from random values, but it took more training iterations. We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, \todo{until reaching a plateau in reconstruction error, at which point we would iterate further using multiplicative updates.} The multiplicative update rules are appealing because they are convex, do not require many choices, and they can be efficiently iterated on a graphics card.\\

There are very few requirements of the theoretical spectral grid. There are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing between points. The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to fit. This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of A- to M-type stars was also sufficiently flexible to model many O-type stars, but there was structured residuals in the wings of hydrogen lines. For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\

We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model. This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \texttt{ATLAS9} code with \project{MARCS} spherical and plane-parallel models atmospheres. While the grid does include spectra for hotter stars computed with ATLAS atmospheres, we restricted ourselves to MARCS model atmospheres for this work. The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum. We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$, truncated the convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements. The matrix $\vec{S}$ includes all \todo{X} convolved BOSZ spectra, with \todo{N} pixels each, which we randomly shuffled at every iteration during factorization. We factorized this matrix into $K = 16$ basis vectors by coordinate descent for \todo{10,000} steps, \todo{and then X steps using multiplicative updates}. After training, the Frobenius loss, also known as the L2 reconstruction loss
\begin{eqnarray}
    L_2 = \norm{{\left(-\log{\vec{S}}\right) - \vec{W}_\star\transpose\vec{F}}}^2
\end{eqnarray}
\noindent{}was found to be \todo{X}. Figure~\ref{fig:schematic} shows some example basis vectors computed by this factorization, where it is somewhat striking that the basis vectors are identifiable by astrophysical mechanism, since no stellar parameters (or astrophysics of any kind) are included in the factorization process. For example, one basis vector just includes hydrogen lines. Another basis vector only has CO-molecule lines. We expand on this in later sections.\\

@@

Telluric transmission can be included in a similar way to stellar absorption. We used a grid of model telluric transmission spectra computed for the \texttt{PypeIt} project. These spectra are computed using the Line-by-Line Radiative transfer Model for an array of observatory locations (latitude, longitude) and atmospheric temperatures, pressures, humidity, and airmass. The original grid of telluric transmission spectra was not available to us. Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, where the grid is compressed using Principal Component Analysis and an inverse hyperbolic sine transformation. We reconstructed the original grid from these approximations, and factorized the negative log of theoretical telluric absorption
\begin{eqnarray}
    -\log\vec{M} \approx \vec{W}\transpose_\earth\vec{G}
\end{eqnarray}
\noindent{}into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$ using $L = 4$ basis vectors. We took \todo{10,000} iterations using coordinate descent, at which point the reconstruction loss was \todo{X}. With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
\begin{eqnarray}
    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta
\end{eqnarray}
where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} in to be solved at inference time.
%We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}).
Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\

@@

\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model in equation \ref{eq:log_y} for the log-transformed data $\vec{Y}$,
\begin{equation}
    \vec{Y} = -\vec{F}\transpose\vecalpha - \vec{G}\transpose\vecbeta + \vec{H}\vecgamma
\end{equation}
\noindent{}which we can compactly write as
\begin{equation}
    \vec{Y} = \vec{A}\vec{X}
\end{equation}
where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
\begin{eqnarray}
    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\-\vec{G}\transpose\\\vec{H}\end{bmatrix}
    \quad \mbox{and} \quad
    \vec{X} = \begin{bmatrix}\vecalpha\\\vecbeta\\\vecgamma\end{bmatrix} \quad .
\end{eqnarray}

@@

We queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from B- to M-type. This selection is not intended to be complete, but is meant to demonstrate the effectiveness of a single linear model for fitting stars across a range of stellar parameters and evolutionary states. For each example star we downloaded a pseudo-random exposure, the details of which are shown in Table~\ref{tab:stars}. No explicit cut was made to ensure a minimum S/N ratio, but in practice most exposures are of high quality. The B- and A-type example stars show significant rotational broadening. For these stars, we adopt the $v\sin{i}$ values reported by \citet{Someone} and convolve the stellar basis vectors before inference. In Section~\ref{sec:discussion} we discuss situations when radial velocity or rotational broadening measurements are not available.\\

@@

The final product of the \harps\ data reduction pipeline is a one-dimensional unrectified spectrum, resampled onto a single wavelength array where all echelle orders have been stitched together. Because we are interested in fitting the continuum on a per-order basis, we instead use the \texttt{e2ds} data products. These are extracted one-dimensional unrectified spectra of each echelle order, with the wavelength solution represented by a low order polynomial. In early experiments we fit the spectra in the \texttt{e2ds} data products directly, an example of which can be seen in Figure~\ref{fig:harps-alf-cen-a}. However, for some exposures we noticed systematic residuals that could be explained by a 2D artefact in the image plane (e.g., a ghost or reflection). While we expect that flat-fielding should have already removed these kinds of artefacts, it did prompt us to explore using the nightly blaze calibration files. Using the calibration blaze files mitigated much of the visible structure, and expectedly provided spectra with a simpler, flatter shape. For these reasons, in all experiments except for the one shown in Figure~\ref{fig:harps-alf-cen-a}, we chose to fit blaze-corrected echelle orders.\\

When fitting each exposure we took the computed Barycentric Earth radial velocity and the radial velocity measured by the \harps\ data reduction pipeline to shift the stellar absorption basis vectors $\vec{F}$ to the observed frame. In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citep{Someone}. Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and \todo{9} Fourier modes per echelle order, totalling \todo{668} model parameters per spectrum.\\

We used the same model setup for the second experiment, where we required many spectra taken of the same star at low and high (e.g., 10 to 200) S/N ratios. The lower bound on S/N ratio might be unrealistic since spectroscopists prefer measuring stellar parameters and chemical abundances from higher quality spectra, but the bound is necessary to evaluate performance at low S/N ratios. The \eso/\harps\ catalog revealed that not many stars meet our repeated S/N sampling requirement. When sorted by number of exposures in descending order, $\alpha$-Centauri A was the first star that met our requirements in repeated S/N sampling. We constructed bins of S/N from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum (Table~\ref{tab:obs-alf-cen-a}) for every S/N ratio bin interval to provide an even sampling of spectra at low and high S/N ratios. We fit our model to each spectrum, treating it as totally indepdendent from all others.\\

\begin{figure*}
    \includegraphics[width=\textwidth, height=\textwidth]{../code/harps_examples.pdf}
    \caption{Continuum-rectified \harps\ spectra of example spectral types from B to M. The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission. \todo{Only BAF-type stars used blaze-corrected spectra. Need to download blazes for others and re-run.}\label{fig:harps-examples}}
\end{figure*}

@@

The first defined experiment involves fitting stars of a variety of spectral types. The rectified spectra for selected example stars in this experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type listed in \project{SIMBAD}. From these rectified spectra it is clear that the model performs well on a variety of stellar types and evolutionary stages. HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16{,}750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorized using NMF. The fit to HD 45418 is quite good, despite this star being well outside the model grid. All other example stars shown in Figure~\ref{fig:harps-examples} have literature stellar parameters that place them well within the bounds of the factorized model grid. Most stars observed with \eso/\harps\ are predictably on the main-sequence, since they are more frequently subject to radial velocity exoplanet searches than evolved stars. HD 203050 (K2) is the only red giant branch in our sample. This is not obvious from Figure~\ref{fig:harps-examples}, suggesting the model does just as well on red giant branch stars as it does on main-sequence stars. This has been our experience in other works using this method on SDSS-V optical and near-infrared spectra (Casey et al., in prep).\\

The cooler stars in the example set have almost no so-called `continuum pixels' blueward of 4000\,\AA. This is expected; the continuum estimate provided by the model matches theoretical rectified spectra. It illustrates that the continuum estimate that we fit simultaneously with stellar absorption does provide a closer estimate to the true continuum rather than a pseudo-continuum.\\

There is good agreement between the model fits and data for example stars of different spectral types. In detail there are mis-matches in line absorption strength, or mis-matches in the form of missing lines (in either the model or the data). We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both. In any case, the model appears to be sufficiently accurate to describe the continuum without substantial bias, so it is not a major concern for our purposes.\\

A separate reason for model-data disagreement comes from our use of the telluric transmission model. The telluric models available to us are of lower spectral resolution ($\mathcal{R} \sim 25{,}000$) than \harps. This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:alf-cen-A-example}), but it is otherwise not obvious. The model telluric transmission predicts features that are clearly broader than the data, but the telluric basis vectors are a sufficiently good description for the data that the telluric absorption does not get captured by the continuum basis vectors and adversely affect the fit.\\

The second defined experiment tests how consistently we can predict continuum for the same star, as a function of S/N ratio. Given an imperfect (or no) model for stellar spectra, noisy signals can mimic stellar absorption, which causes the continuum to become systematically biased at low S/N ratios. We fit \todo{X} spectra of $\alpha$-Centauri A, which are sampled uniformly in S/N bins from 10 to 200 pixel$^{-1}$. This experiment forces us to select a summary statistic to evaluate the consistency of our continuum rectification. We adopt the 95th percentile in rectified flux as a statistic to compare to normalisation methods described in the literature, but we prefer to view how the \emph{distribution} of normalized flux values change as a function of S/N ratio. Both properties are shown in Figure~\ref{fig:repeatability-with-snr}, where the line indicates the 95th percentile of normalized flux, and the shaded area shows the distribution of normalized flux values. Above S/N $\sim$ 30, both indicators are extremely stable and do not vary significantly with S/N ratio. This is notable, as most spectra acquired for the purpose of stellar parameters and chemical abundances will aim for S/N $>$ 30, but spectra below S/N $<$ 50 are routinely described as being problematic to rectify consistently. From this figure it highlights that we can achieve consistent continuum normalisation whether the star has S/N $\sim$ 30 or 300. \todo{This experiment was done on non-blaze-corrected spectra. I either need to re-do it, or update the text accordingly.}\\

We can make quantitative statements on the consistency of continuum rectification for $\alpha$-Centauri A. Taking the 95th percentile of normalized flux as a proxy statistic for the normalization consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalized flux is 0.22\%, which is equivalent a spread in normalized flux of 0.0022. This variation is smaller than the typical flux errors, and only increases marginally when we include lower quality spectra. Taking all spectra with S/N $>$ 50, we find the scatter in the 95th percentile of normalized flux to be 0.35\%. With S/N $>$ 30, this rises marginally to 0.46\%. In all scenarios the estimated error in continuum is less than the typical flux error.\\
